### 2021-07-04 rad-icfp-2021/purerad.tex
rewrote contribution
rewriting section 3

\subsection{Source Language}

@@

We consider a standard call-by-value language. 
It consist in a first-order functional language enriched with arrays and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.

@@

The typing rules are in Figure~\ref{fig:source_typesystem}. We have included a minimal set of array operations for the sake of illustration, 
but it is not difficult to add further operations. 
For scalar operations, we assume given a set of operations, including $+$ and $*$. 
!op1! and !op2! denote respectively a unary and a binary smooth operation on reals. 
Typical examples include !cos, exp, +, *!. 
We use infix notation for binary operators.
For the sake of simplicity in the presentation, a bound function in an array operation !map, map2, foldl! is restricted to having no free variables.
For the same reason, we restrict arrays of reals.
The general cases are presented in Section~\ref{sec:generalisation}.

@@

The target language for differentiation is an extension to the source language, 
which is higher-order as needed to carry the continuation used to have a pure reverse-mode transformation, 
and contains more array primitives. In addition, we require the set of operations to be closed under partial differentiation.

@@

In more detail, for every unary scalar operation !op1!, 
we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
representing respectively the first and second partial derivative of !op2!.
Similarly, some additional operations on arrays are needed. 
!scanl! is a !foldl! that stores all intermediate results. 
!scanr! is similar, but reads the array from right to left, and also stores from right to left.
!reduce! is a !foldl! for which the function is associative. 
This means in practice that it can be computed faster.
We also include a !map3! operator. 
Though not necessary as it can be simulated with two !map2!, it is a convenient notation and can sometimes run faster in practice.
Finally, we add a somewhat strange operator which we call !shift1!. 
It takes an array of size $n$,  shifts all the element of the array by one, 
forgetting the last element, and puts a one in the first place. 
This operator naturally shows up when differentiating !foldl!.

The type system for the extended grammar of target is presented in Figure~\ref{fig:target_typesystem}.

@@

A standard semantics of this language is given in Section~\ref{sec:correctness}.

@@

In Figure~\ref{fig:direct_diff_macro} we present the direct transformation for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
This transformation may be hard to read and prove correct. 
In Section~\ref{sec:unf} we decompose this transformation into several easier 
and more intuitive steps via a new intermediate representation which we call UNF.

>>>>

\subsection{Source Language}
\label{sub:sourcelang}

@@

We consider a standard call-by-value language. 
It consists of a first-order functional language with arrays and a few typical second-order array operations. 
The types !T1,T2! and terms !e1,e2! are given in Figure~\ref{fig:source_grammar}.
We have included a minimal set of array operations for the sake of illustration,  it is not hard to add more.
See Section~\ref{sec:generalisation}.

@@

The typing rules are given in Figure~\ref{fig:source_typesystem}. 
For scalar operations, we assume given a set of operations, including $+$ and $*$. 
!op1! and !op2! denote respectively a unary and a binary operation on reals. 
These operations represent smooth total functions, 
but again this can be easily generalised (\S\ref{sec:generalisation}).  
Typical examples include !cos, exp, +, *!. 
We use infix notation for binary operators.

For the sake of simplicity in the presentation, the bound function in !reduce! 
is restricted to having no free variables. For the same reason, we currently restrict to arrays of reals.
We show how to lift these restrictions in Section~\ref{sec:generalisation}.

@@

The target language of our source-code transformation is an extension to the source language.
It is a higher-order language, as our purely functional reverse-mode introduces a continuation.
The set of scalar operations should also be closed under partial differentiation. 
In more detail, for every unary scalar operation !op1!, 
we assumed given an operator $\partial_1$!op1! whose semantics should be the derivative of !op1!.
For every binary operator !op2!, we assume given operators $\partial_1$!op2!, $\partial_2$!op2! 
representing respectively the first and second partial derivative of !op2!.

TODO:redo
Similarly, the target language contains more array primitives which are used to define the reverse derivatives of array operations. 
!scanl! is a !foldl! that stores all intermediate results. 
!scanr! is similar, but reads the array from right to left, and also stores from right to left.
!reduce! is a !foldl! for which the function is associative. 
This means in practice that it can be computed faster.
Finally, we add a somewhat strange operator which we call !shift1!. 
It takes an array of size $n$,  shifts all the element of the array by one, 
forgetting the last element, and puts a one in the first place. 
This operator naturally shows up when differentiating !reduce!.

@@

The type system for the extended grammar of the target language is presented in Figure~\ref{fig:target_typesystem}.

@@

@@

TODO:redo
In Figure~\ref{fig:direct_diff_macro} we present the direct transformation for pure reverse mode differentiation.
We explain in the next Section (\ref{sub:Partial evaluation and optimisation}) how to optimise it further and compute the gradient.
This transformation may be hard to read and prove correct. 
In Section~\ref{sec:unf} we decompose this transformation into several easier 
and more intuitive steps via a new intermediate representation which we call UNF.
